---
title: "Nonparametric MLR - Lab Activity - Example Solution"
author: "B.A. Hartlaub and A.S. Wagaman"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r setup, include = FALSE} 
library(NSM3)
library(tidyverse)
library(mosaic) 
library(SuppDists)
library(Rfit)
```


# Nonparametric Regression Examples 

## Example 1 (includes parametric regression comparison)

We use a data set on painted turtles, where we know their biological sex and their height/length/width measurements. 

```{r}
turtle <- 
  read.table("https://awagaman.people.amherst.edu/nonparametricstats/paintedturtle.txt", 
      header = TRUE)
```

Suppose we want to try to predict the height of the turtles, using length and sex. 

Fit a nonparametric regression predicting height using length. Use *theil* for exact results since this is an SLR. Do your results make sense?

> SOLUTION:

```{r, eval = FALSE}
# yikes this doesn't run!
# with(turtle, theil(length, height, beta.0 = 0, type = "t"))
# try without the plot
with(turtle, theil(length, height, beta.0 = 0, type = "t", doplot = F))
```

We get output with the second command (the first was apparently interrupted because of issues plotting), however, the `beta.hat` and `alpha.hat` estimates are both NA. We got a CI for beta though, and there's clearly a significant relationship because `C.bar` is 0.829 and the corresponding p-value is rounded to 0.

You probably noticed something odd going on. There are many ties here among the length values. It appears that these are causing issues. If we want the nonparametric estimates, we'll have to get them ourselves.

How would we do that? A little brute code (which can likely be done more elegantly). Just follow along with what is being done here. 

```{r}
num <- nrow(turtle)
slopes <- NULL
for(i in 1:(num-1)){
  for(j in (i+1):num){
     slopes <- c(slopes, as.numeric(lm(height ~ length, 
                                       data = turtle[c(i,j),])$coeff[2]))    
  }
} 
favstats(~ slopes)
gf_boxplot(~ slopes)
```

This is an approach where we iterate and each time, add a possible slope to our slopes vector. We use the parametric lm function - you could compute this with the respective X and Y's too. Why do we pull the second entry from the coefficients? 

> SOLUTION:

In an lm, the first coefficient is the intercept. The second is the slope coefficient for the first term. Since this is SLR, that's the one we want. 


What would you report as the nonparametric slope estimate?

> SOLUTION: 

We saved the possible slopes into a vector, and we want the median of the possible slopes as our estimate. Thus, we'd report the median of 0.382ish as our slope estimate (depending on where you round). 

We could then get the alpha (intercept estimate) fairly easily as well.

```{r}
alphas <- with(turtle, height - 0.382*length)
median(alphas)
```

The nonparametric regression line thus obtained manually would be:

> SOLUTION: 

$$
Predicted Median Height= -1.419 + 0.382*length
$$


Compare your results to those from `rfit` rank-based approach for the SLR. How much do the lines differ? 

> SOLUTION: 

```{r}
myht <- rfit(height ~ length, data = turtle)
summary(myht)
```

The rfit approach yields a predicted line of:

$$
PredictedMedianHeight = -2.435 + 0.391 * length.
$$

The slopes are very similar, as we obtained 0.382 and 0.391, while the intercepts differ a little bit more. 

Now we want to extend to multiple linear regression. 

Use `sen.adichie` to test whether or not the slopes of the regression line for predicting height using length would be equal or different for the two sexes.  Remember a particular data format is needed for this procedure.

> SOLUTION: 

To begin, although this was not requested, we'll plot this to use our intuition about what should happen. You can add the parametric fitted lines or not.

```{r}
ggplot(turtle, aes(x = length, y = height, color = sex)) + 
  geom_point() + geom_lm() # assumes slopes can differ
```

To run `sen.adichie`, we have to get the data in the correct format - separate data files for each group, with only the x and y (and in order, first x, then y), compiled and put into a list. 

```{r}
mturtle <- filter(turtle, sex == "male") %>%
  dplyr::select(length, height)
fturtle <- filter(turtle, sex == "female") %>%
  dplyr::select(length, height)
z <- list(mturtle, fturtle)
sen.adichie(z)
```

Here, our V is 9.794 and the corresponding p-value is 0.002. We would reject the null hypothesis at the commonly used significance levels. This indicates that we need non-parallel slopes.  That is, a model where the groups should be able to have different slopes would be better. 


Try `rfit` with a categorical predictor and interaction term to do the same test as `sen.adichie`. Are the results consistent? 

> SOLUTION:

```{r}
# remember if K > 2, this means fitting two models, then using drop.test
# if K = 2, there is another option

modint <- rfit(height ~ length*sex, data = turtle)
modmain <- rfit(height ~ length + sex, data = turtle)

# first approach, fit interaction model, check single interaction slope
summary(modint)

# second approach, use drop.test to compare models
drop.test(modint, modmain)
```

In the first approach (which relies on K being 2), we fit a model with an interaction term between length and sex to see if we have the same results as `sen.adichie`. If the interaction term is significantly different from 0, then the two groups require different slopes for length. We can see the p-value for the interaction term is 0.01739, which is well below 0.05. Thus, we reach the same conclusion as `sen.adichie` - we should not use parallel slopes here.

In the second approach, we use a `drop.test()` to compare nonparametric models - you compare the model with and without the interaction. The p-value is again small (0.0079), so we reach the same conclusion with this approach. 


Try `lm` following the same approach as above. Are the nonparametric results about whether you need different slopes for the sexes consistent with the parametric results?

> SOLUTION:

```{r}
# remember if K > 2, this means fitting two models, then using anova
# if K = 2, there is another option
  
paramodint <- lm(height ~ length * sex, data = turtle)
paramodmain <- lm(height ~ length + sex, data = turtle)

# first approach, fit interaction model, check single interaction slope
msummary(paramodint)

# second approach, use anova to compare models, swap order!
anova(paramodmain, paramodint)
```

Under either approach, we see tiny p-values for the interaction coefficient / corresponding test. Thus, the parametric approach agrees with the nonparametric ones. We should have different slopes for the sexes here for length when trying to predict height. 

## Example 2

```{r}
primate <- 
  read.table("https://awagaman.people.amherst.edu/nonparametricstats/primatedata.txt", 
                      header = TRUE)
```

The data set *primatedata.txt* contains variables about shoulder blades from 105 primates. There are 5 indices (basically measured distances): AD.BD, AD.CD, EA.CD, Dx.CD, and SH.ACR, and 2 angles recorded (EAD and beta) for each shoulder blade. The primate genus (plural: genera) is also recorded, as the variable *class*. Gibbons (Hylobates), orangutangs (Pongo), chimpanzees (Pan), gorillas (Gorilla), and man (Homo) are all represented in the data set.

Pick three of the five genera to focus on. Which three do you think it would be interesting to explore differences between?

> SOLUTION:

Any three genera can be picked. Later answers may differ based on the three genera you picked. 

For the purpose of this solution, we decided to pick Homo, Pan, and Pongo. 

Create a data set with only your three genera present. It may help to add a mutate where you reset the factor levels (just do: `class = factor(class)`) so it understands there are only 3 levels. 

> SOLUTION:

```{r}
mythree <- filter(primate, class == "Pan" | class == "Pongo" | class == "Homo") %>%
  mutate(class = factor(class))
```

Use `sen.adichie` to test whether or not the slopes of the regression line for predicting AD.BD using AD.CD would be equal or different for your choice of three genera.  

> SOLUTION: 

```{r}
Homo <- filter(mythree, class == "Homo") %>%
  dplyr::select(AD.CD, AD.BD)
Pan <- filter(mythree, class == "Pan") %>% 
  dplyr::select(AD.CD, AD.BD)
Pongo <- filter(mythree, class == "Pongo") %>% 
  dplyr::select(AD.CD, AD.BD)
z <- list(Homo, Pan, Pongo)
sen.adichie(z)
```


Our test statistic V is 21.43 and the p-value is reported as 0. Based on our evidence, the slopes do not appear to be equal for the group of three genera selected. Thus, we would want to use an interaction term with class.


Use `rfit` to test whether or not you can predict *EAD* using all 5 indices (AD.BD, AD.CD, EA.CD, Dx.CD, and SH.ACR) with each (separately) interacted with genera (which includes genera's main effect), versus a null model.  

> SOLUTION:

This is intended to be on your subset of three genera (so the output isn't as long). 

```{r}
eadmodfull <- rfit(EAD ~ AD.BD*class + AD.CD*class + EA.CD*class +
                    Dx.CD*class + SH.ACR*class, data = mythree)
summary(eadmodfull, overall.test = 'drop') # summary(eadmodfull) works too
```

The test statistic is 16.65 with a corresponding p-value of basically 0. There is evidence at least one term in the model (and there are many!) is useful for predicting EAD. 

Use `rfit` to test whether or not you can drop ALL the interactions from the model above and still predict *EAD* with the 5 indices and genera as main effects. 

> SOLUTION:

```{r}
eadmodreduced <- rfit(EAD ~ AD.BD + AD.CD + EA.CD + Dx.CD + SH.ACR + class, 
                      data = mythree)
drop.test(eadmodfull, eadmodreduced) # larger model first
```

Yay! The p-value here is large. This suggests that we can actually drop all the interaction terms from the model. If you want to compare model summaries, you can. We only have a drop in robust $R^2$ of about 0.03, and we got rid of ALL the interaction terms. It's a much simpler model. 

```{r}
summary(eadmodreduced)
```


(If time permits) Find the "best" nonparametric model you can to predict beta, using any combination of predictors, interactions, re-expressed predictors, etc. for your data set with only 3 genera. Remember, we like simple models for ease of interpretation. (If you don't care about interpreting, then you can make a model more complicated if it improved prediction.) Your response should use at least 2 nonparametric test procedures (though they don't need to be drop-tests), and be sure you fit nonparametric models. Feel free to make comparisons to parametric methods. 

> SOLUTION:

Answers will vary. The idea is to explore using the drop.tests for subsets of predictors, and individual tests for individual terms in the presence of the others. The overall.test could be used to test for whether any term is useful. 


## Example 3 - Simulated Data

Here, we consider a setup with simulated data with the SAME underlying linear model but different distributions for the error terms.  We generated the data separately and have uploaded the data sets. We are not sharing the linear model with you - see if you can figure it out while exploring the problems below.

The setting is as follows: The company WidgetsRUs sells widgets. They have three stores in a particular city, and they are trying to develop a model to predict the daily number of widgets sold based on some other variables: price of the widgets that day (dollars), whether there was a publicized sale on widgets that day (yes/no), and the location of the store (3 locations). They have provided data for you from the past 100 days of sales, which you can assume is a representative sample of their sales. 

### Normal Errors

```{r}
WidgetNormal<- 
  read.csv("https://awagaman.people.amherst.edu/nonparametricstats/WidgetDataNormal.csv",
           h = T)
str(WidgetNormal)
```

Find a "good" parametric regression line for predicting the daily NumSold values based on the other predictors. (No, you do not need to use all the predictors).

> SOLUTION: 

```{r}
# plot parametric solutions to explore
ggplot(WidgetNormal, aes(x = Price, y = NumSold, color = Location, shape = Sale)) +
  geom_point() + geom_lm()

# fit the largest model we can - it has a three way interaction!
parafull <- lm(NumSold ~ Price*Location*Sale, data = WidgetNormal)
msummary(parafull)

# the three-way interaction is "scary"; maybe we don't need it
para2way <- lm(NumSold ~ Price*Location + Price*Sale + Location*Sale, 
               data = WidgetNormal)
anova(para2way, parafull)

msummary(para2way) # several other interactions aren't significant
# we will try dropping some

# tried this model next
paralm <- lm(NumSold ~ Price*Location + Sale, data = WidgetNormal)
anova(paralm, para2way)

# sale isn't significant so take it out (overwrote model)
msummary(paralm)

paralm <- lm(NumSold ~ Price*Location, data = WidgetNormal)
msummary(paralm)

# could we really drop all that?
anova(paralm, parafull)
```

After following a process to try to simplify the model, we ended up with a reduced model with just *Price* interacted with *Location*. The nested F-test suggests that we can go with the smaller model, which has an $R^2$ of 0.8445, which is not much loss (at all, compared to 0.8479 for the full model). In fact, the adjusted $R^2$ values are almost the same, 0.8421 for the full model and 0.8419 for the smaller model.  Thus, we will use the smaller parametric model here. 

Fit a nonparametric competitor using the same model structure. 

> SOLUTION:

```{r}
nplm <- rfit(NumSold ~ Price*Location, data = WidgetNormal)
summary(nplm, overall.test = 'drop')
```


How do the models compare in terms of coefficient estimates? In terms of model fit? In terms of satisfying conditions? (You can get the residuals from the nonparametric models and make plots manually.)

> SOLUTION:

```{r}
paralm$coeff
nplm$coeff
```

The coefficients look to be very similar. 

The parametric $R^2$ is 0.8445 and the nonparametric robust $R^2$ is 0.752. While the nonparametric $R^2$ is a little lower than the parametric $R^2$, both indicate a very good fit. 

For conditions, we make plots! We assume the linear model is appropriate, that we have a random sample of errors which are independent from each other, and with mean/median 0. 

For the parametric model:

```{r}
mplot(paralm, which = 1)
mplot(paralm, which = 2)
```

The constant variance condition seems reasonable, though the variance may decrease slightly in the middle and towards the end. A few outliers are labeled. The QQ plot seems to indicate reasonable normality as well, a minor deviation in the upper tail corrects itself. This may be due to the discrete nature of the response variable. 

For the nonparametric model, we have the residuals, but the plot functionality is not automated. We can make the plots manually though. We need to check for similar spread still, and instead of normality, just need a continuous population. 

```{r}
gf_dens(~ nplm$residuals) # looks somewhat normal 

# make residuals versus fitted plot manually
gf_point(residuals(nplm) ~ fitted(nplm), 
         xlab = "Predicted Values", ylab = "Residuals")
```

The constant variance condition is the one that likely still has the concern. We can check across the locations to see if there is a huge difference in spread. (Like you would check for an ANOVA).

```{r}
favstats(residuals(nplm) ~ Location, data = WidgetNormal)
gf_boxplot(residuals(nplm) ~ Location, data = WidgetNormal)
```

The spreads do not seem extremely different by location at least.  (Note that since we generated them from the same population, technically, the simulation was set up with the condition met.)

### Double Exp Errors

```{r}
WidgetDExp <- 
  read.csv("https://awagaman.people.amherst.edu/nonparametricstats/WidgetDataDExp.csv", 
                       h = T)
```

Find a "good" parametric regression line for predicting the daily NumSold values based on the other predictors. (No, you do not need to use all the predictors).

> SOLUTION: 

```{r}
parafull <- lm(NumSold ~ Price*Location*Sale, data = WidgetDExp)
msummary(parafull)
# does the model from above still work?
paralm <- lm(NumSold ~ Price*Location, data = WidgetDExp)
msummary(paralm)
anova(paralm, parafull)
```

Following the same process as with the normal errors, we find a parametric nested F-test suggests that we can go with the smaller model with just *Price***Location*, which has an $R^2$ of 0.58, which is not much loss (at all) from the full $R^2$ of 0.5866. In fact, the smaller model has a higher adjust $R^2$ so we will use it here. We do note that this is a major drop in $R^2$ compared to when the errors are normal. The model just fits moderately well now. 


Fit a nonparametric competitor using the same model structure. 

> SOLUTION:

```{r}
nplm <- rfit(NumSold ~ Price*Location, data = WidgetDExp)
summary(nplm, overall.test = 'drop')
```

How do the models compare in terms of coefficient estimates? In terms of model fit? In terms of satisfying conditions? (You can get the residuals from the nonparametric models and make plots manually.)

> SOLUTION:

```{r}
paralm$coeff
nplm$coeff
```

These coefficients are still very similar to each other. 

The parametric $R^2$ is 0.58 and the nonparametric robust $R^2$ is 0.568. While the nonparametric $R^2$ is a little lower than the parametric $R^2$, it is closer than it was with the normal errors. 

As before, for conditions, we make plots! We assume the linear model is appropriate, that we have a random sample of errors which are independent from each other, and with mean/median 0. 
For the parametric model:

```{r}
mplot(paralm, which = 1)
mplot(paralm, which = 2)
```

The constant variance condition seems reasonable, better than before, in fact. The QQ plot seems to indicate minor issues with normality with deviations from linear trend in both tails. The length of each tail (it's not just like 3 points) suggests this is likely serious enough to say the condition is not met.  We could try a re-expression to see if that helps, if desired.

For the nonparametric model, we make the plots manually. We need to check for similar spread still, and instead of normality, just need a continuous population. 

```{r}
# looks bell-shaped, but likely has heavier tails than a normal
gf_dens(~ nplm$residuals) 
gf_point(residuals(nplm) ~ fitted(nplm), 
         xlab = "Predicted values", ylab = "Residuals")
```

The constant variance condition is reasonable (the decrease in the middle is due to the sale/no sale transition, perhaps). We can check across the locations to see if there is a huge difference in spread. (Like you would check for an ANOVA).

```{r}
favstats(residuals(nplm) ~ Location, data = WidgetDExp)
gf_boxplot(residuals(nplm) ~ Location, data = WidgetDExp)
```

The SDs of the residuals across *Location* differ a bit more than we might like. However, (4.2/3.27) is max/min ratio of SDs, and that's below 2, and the IQRs in the boxplot are similar. (Note that since we generated them from the same population, technically, the simulation was set up with the condition met.)


Are you able to find a nonparametric competitor (with different model structure) that does better than this in terms of model fit or in terms of satisfying conditions?

> SOLUTION: 

```{r}
nplm2 <- rfit(NumSold ~ Price*Location*Sale, data = WidgetDExp)
summary(nplm2, overall.test = 'drop')
```

The full model with three-way interactions only has a robust $R^2$ of 0.576. We are within 1% of that so that is pretty good in terms of $R^2$. Maybe we can improve in simplicity. Here, a lot of predictors are not significant given the presence of others in the model. Let's see how the nonparametric SLR does.

```{r}
nplm3 <- rfit(NumSold ~ Price, data = WidgetDExp)
summary(nplm3, overall.test = 'drop')
drop.test(nplm2, nplm) # can use just Price*Location
drop.test(nplm, nplm3) # want to keep the interaction and/or Location
```

It seems our model is pretty solid. 

#### Cauchy Errors

Note to avoid many negative *NumSold* values, we had to use a very small spread value for the Cauchy errors compared to the Normal and DExp cases. Two values were forced positive by redrawing errors. 

```{r}
WidgetCauchy <- 
  read.csv("https://awagaman.people.amherst.edu/nonparametricstats/WidgetDataCauchy.csv", 
                         h = T)
```

Find a "good" parametric regression line for predicting the daily *NumSold* values based on the other predictors. (No, you do not need to use all the predictors).

> SOLUTION: 

```{r}
gf_point(NumSold ~ Price, data = WidgetCauchy)
parafull <- lm(NumSold ~ Price*Location*Sale, data = WidgetCauchy)
msummary(parafull)
paralm <- lm(NumSold ~ Price*Location, data = WidgetCauchy)
msummary(paralm)
anova(paralm, parafull)
```

Following the same process as with the normal errors, we find a parametric nested F-test suggests that we can go with the smaller model with just *Price***Location*, which has an $R^2$ of 0.2509, which is not much loss from the full $R^2$ of 0.2614. Thus, we will use the smaller model here. We do note that this is a major drop in $R^2$ compared to when the errors are normal, or when the errors were double exponential. The model does not fit very well now. We can see why in the plot - there are extreme outliers. 

Fit a nonparametric competitor using the same model structure. 

> SOLUTION:

```{r}
nplm <- rfit(NumSold ~ Price*Location, data = WidgetCauchy)
summary(nplm, overall.test = 'drop')
```

Tada! The Robust R-Squared here is 0.8665, which is MUCH higher than the parametric counterpart. (This is why we suggest removing outliers to assess impact on analysis with parametric procedures.)


How do the models compare in terms of coefficient estimates? In terms of model fit? In terms of satisfying conditions? (You can get the residuals from the nonparametric models and make plots manually.)

> SOLUTION:

```{r}
paralm$coeff
nplm$coeff
```

These coefficients are no longer similar. They have the same signs, but there is quite a difference in magnitude. 

The parametric $R^2$ is 0.2509 and the nonparametric robust $R^2$ is 0.8665. The parametric model is a much worse fit than the nonparametric model here.  

As before, for conditions, we make plots! We assume the linear model is appropriate, that we have a random sample of errors which are independent from each other, and with mean/median 0. 
For the parametric model:

```{r}
mplot(paralm, which = 1)
mplot(paralm, which = 2)
```

Nope! The outliers cause issues with constant variance and the heavy tails mean the distribution does not appear to be normal. 

For the nonparametric model, we make the plots manually. We need to check for similar spread still, and instead of normality, just need a continuous population. 

```{r}
gf_dens(~ nplm$residuals) # one peak but has a long + tail
gf_point(residuals(nplm) ~ fitted(nplm), 
         xlab = "Predicted values", ylab = "Residuals")
```

The constant variance condition has issues with the outliers, and the distribution of residuals has a long tail. Luckily, the outliers did not influence the regression line as much as they did in the parametric case. We can check across the locations to see if there is a huge difference in spread. (Like you would check for an ANOVA).

```{r}
favstats(residuals(nplm) ~ Location, data = WidgetCauchy)
gf_boxplot(residuals(nplm) ~ Location, data = WidgetCauchy)
```

The SDs differ a bit more than we might like due to the extreme outliers from the Cauchy, but the IQRs in the boxplot and seen in favstats are similar.

(Note that since we generated them from the same population, technically, the simulation was set up with the condition met.)

Are you able to find a nonparametric competitor (with different model structure) that does better than this in terms of model fit or in terms of satisfying conditions?

> SOLUTION: 

The results are very similar to what we found for the last type of errors.

```{r}
nplm2 <- rfit(NumSold ~ Price*Location*Sale, data = WidgetCauchy)
summary(nplm2, overall.test = 'drop')
```

The full model with three-way interactions only has a robust $R^2$ of 0.8669. We are within 0.0004 of that so we are doing very well in terms of $R^2$. As before, maybe we can improve in simplicity. Here, a lot of predictors are not significant given the presence of others in the model. Let's see how the SLR does.

```{r}
nplm3 <- rfit(NumSold ~ Price, data = WidgetCauchy)
summary(nplm3, overall.test = 'drop')
drop.test(nplm2, nplm) # can use just Price*Location
drop.test(nplm, nplm3) # want to keep the interaction and/or Location
```

The SLR actually has a pretty big drop in terms of $R^2$.  The drop in dispersion tests confirm our findings. It seems our model is pretty solid. 

#### Model Comparison

The underlying model structure did NOT change in any of the settings above. The errors changed in distribution. 

Were the parametric coefficient estimates "stable"?

> SOLUTION: 

We saved over the names, so here are the three models and their coefficients (just the coeffs). 
```{r}
paralmN <- lm(NumSold ~ Price*Location, data = WidgetNormal)
paralmD <- lm(NumSold ~ Price*Location, data = WidgetDExp)
paralmC <- lm(NumSold ~ Price*Location, data = WidgetCauchy)
coefficients(paralmN)
coefficients(paralmD)
coefficients(paralmC)
```

We see that the slope estimates are fairly different, and the location effects are also fairly radically changing. None of them change sign, but we'd definitely be reporting different lines with each of these models. 

Were the nonparametric coefficient estimates "stable"?

> SOLUTION:

We saved over the names, so here are the three models and their coefficients (just the coeffs). 

```{r}
nplmN <- rfit(NumSold ~ Price*Location, data = WidgetNormal)
nplmD <- rfit(NumSold ~ Price*Location, data = WidgetDExp)
nplmC <- rfit(NumSold ~ Price*Location, data = WidgetCauchy)
coefficients(nplmN)
coefficients(nplmD)
coefficients(nplmC)
```

The DExp case seems to have thrown the nonparametric estimators a bit, though they aren't as far off as the parametric estimates in that setting. But the Cauchy and Normal nonparametric estimates are very close to each other. 

Is there anything else you learned from these comparisons?

> SOLUTION:

Answers will vary. 

What we learned setting this up for you explore: Setting up simulated data can be tricky. But the Cauchy data illustrates what we wanted you to see very nicely. 

Note that each of these data sets was a single draw of the errors from their corresponding distributions. To truly assess behavior here, doing this repeatedly would be advised. 

## Data Sources

Painted turtle data from Jolicoeur, Pierre, and James E. Mosimann. "Size and shape variation in the painted turtle. A principal component analysis." Growth 24.4 (1960): 339-354.

Primate data from Izenman, Alan J. Modern multivariate statistical techniques. Vol. 1. New York: Springer, 2008.

Original primate data source: Ashton, E. H., C. E. Oxnard, and T. F. Spence. "Scapular shape and primate classification." Proceedings of the Zoological Society of London. Vol. 145. No. 1. Oxford, UK: Blackwell Publishing Ltd, 1965.

The Widget data was simulated.  

## Standard Reference 

These materials were created for use in an undergraduate nonparametrics course. Where provided, textbook references refer to Nonparametric Statistical Methods (3rd Edition) by Hollander, Wolfe, and Chicken. Notation is chosen to be consistent with that text. However, materials may be used independently of the textbook. 

## License

This work is protected by creative commons license CC BY-NC-SA. 
